\textbf{SPECFEM3D}\\ \textbf{User Manual}

### 2010-11-18 doc/USER_MANUAL/manual_SPECFEM3D.tex
updates SPECFEM3D manual and example readme files; bug fixes coupling surface detection for single layer cases

\author{$\copyright$ California Institute of Technology (USA) and\\
University of Pau / CNRS / INRIA (France)\\
Version 2.0.0}

\maketitle
\newpage{}

@@

The software package SPECFEM3D simulates seismic
wave propagation at the local or regional scale
based upon the spectral-element method (SEM).
The SEM is a continuous Galerkin technique, which can easily be made discontinous \citep{BeMaPa94,KoWoHu02,ChCaVi03,Kop06,WiStBuGh10}; it is then a particular case of the discontinuous Galerkin technique \citep{ReHi73,FaRi99,HuHuRa99,CoKaSh00,GiHeWa02,RiWh03,MoRi05,GrScSc06,BeLaPi06,DuKa06,DeSeWh08,WiStBuGh10,DeSe10,EtChViGl10}, with optimized efficiency because of its tensorized basis functions.
In particular, it can accurately handle very distorted mesh elements \citep{OlSe11}.
Note that in many (most?) geological models in the context of seismic wave propagation studies
(except for fault dynamic rupture studies)
a discontinous mesh is not needed because material property contrasts are not drastic and thus a continuous formulation is sufficient. For a detailed introduction to the SEM as applied to regional
seismic wave propagation, please consult \citet{KoVi98,KoTr99,ChKoViCaVaFe07,TrKoLi08} and
in particular \citet{KoLiTrSuStSh04}.

Effects due to lateral variations in compressional-wave speed, shear-wave
speed, density, a 3D crustal model, topography and bathymetry are
included. The package can accommodate
full 21-parameter anisotropy (see~\citet{ChTr07}) as well as lateral
variations in attenuation. Adjoint capabilities and finite-frequency
kernel simulations are included \citep{LiTr06,TrKoLi08,FiIgBuKe09}.

The SPECFEM3D package was first developed by Dimitri Komatitsch and Jeroen Tromp at Harvard University, then Caltech and Princeton, USA, starting in 1998, based in part on earlier work by Dimitri Komatitsch and Jean-Pierre Vilotte at IPG in Paris, France from 1994 to 1997. Since then it has been developed and maintained by a development team: in alphabetical order,
Emanuele Casarotti,
Min Chen,
Vala Hj\"orleifsd\'ottir,
Emiljana Jorgji,
Sue Kientz,
Dimitri Komatitsch,
Qinya Liu,
Yang Luo,
Nicolas Le Goff,
Pieyre Le Loher,
Alessia Maggi,
Roland Martin,
Dennis McRitchie,
David Mich\'ea,
Daniel Peter,
Brian Savage,
Leif Strand,
Carl Tape,
Jeroen Tromp
\toall{add full list of developers here and update the AUTHORS file accordingly}.

\toall{We should also update the Adobe Illustrator / PDF cover page as well at some point}

All SPECFEM3D\_GLOBE software is written in Fortran90 with full portability
in mind, and conforms strictly to the Fortran95 standard. It uses
no obsolete or obsolescent features of Fortran77. The package uses
parallel programming based upon the Message Passing Interface (MPI)
\citep{GrLuSk94,Pac97}.

@@

If you use SPECFEM3D for your own research, please cite at least one
of the following articles: \cite{TrKoLi08,VaCaSaKoVi99,LeChLiKoHuTr08,LeChKoHuTr09,LeKoHuTr09,KoMiEr09,KoErGoMi10,KoGoErMi10,WiKoScTr04,KoLiTrSuStSh04,ChKoViCaVaFe07,MaKoDi09,KoViCh10,CaKoLaTiMiLeSnTr08,TrKoHjLiZhPeBoMcFrTrHu10,KoRiTr02,KoTr02a,KoTr02b,KoTr99} or \cite{KoVi98}.
If you work on geophysical applications, you may be interested in citing some of these application articles as well, among others: \cite{WiKoScTr04,JiTsKoTr05,KrJiKoTr06a,KrJiKoTr06b,LeChLiKoHuTr08,LeChKoHuTr09,LeKoHuTr09,ChFaKo04,FaChKo04,RiRiKoTrHe02,GoAmTaCaSmSaMaKo09,TrKo00,SaKoTr10}.
The corresponding Bib\TeX{} entries may be found
in file \texttt{USER\_MANUAL/bibliography.bib} or in comments at the
beginning of file \texttt{specfem3D.f90}.

@@

@@

The SPECFEM3D software package relies on the Scotch library to partition the meshes created with CUBIT.
Note that we use CUBIT to create meshes of hexahedra, but other packages can be used as well, for instance GiD from \url{http://gid.cimne.upc.es} or Gmsh from \url{http://geuz.org/gmsh}. Even mesh creation packages that generate tetrahedra, for instance TetGen from \url{http://tetgen.berlios.de}, can be used because each tetrahedron can then easily be decomposed into four hexahedra as shown in the picture of the TetGen logo at \url{http://tetgen.berlios.de/figs/Delaunay-Voronoi-3D.gif}; while this approach does not generate hexahedra of optimal quality, it can ease mesh creation in some situations and it has been shown that the spectral-element method can very accurately handle distorted mesh elements \citep{OlSe11}.

@@

\begin{description}
\item [{\texttt{F90}}] Path to the Fortran90 compiler.
\item [{\texttt{MPIF90}}] Path to MPI Fortran90.
\item [{\texttt{MPI\_FLAGS}}] Some systems require this flag to link to
MPI libraries.
\item [{\texttt{FLAGS\_CHECK}}] Compiler flag for non-critical subroutines.
\item [{\texttt{FLAGS\_NO\_CHECK}}] Compiler flag for creating fast, production-run
code for critical subroutines.
\end{description}
The configuration scripts automatically creates for each executable a corresponding
\texttt{Makefile} in the \texttt{src/} subdirectoy.
The \texttt{Makefile} contains a number of suggested entries for various
compilers, e.g., Portland, Intel, Absoft, NAG, and Lahey. The software
has run on a wide variety of compute platforms, e.g., various PC clusters
and machines from Sun, SGI, IBM, Compaq, and NEC. Select the compiler
you wish to use on your system and choose the related optimization
flags. Note that the default flags in the \texttt{Makefile} are undoubtedly
not optimal for your system, so we encourage you to experiment with
these flags and to solicit advice from your systems administrator.
Selecting the right compiler and optimization flags can make a tremendous
difference in terms of performance. We welcome feedback on your experience
with various compilers and flags.

@@

\begin{description}
\item [{\texttt{CUSTOM\_MPI\_TYPE}}] Set to \texttt{MPI\_REAL} for single
precision and \texttt{MPI\_DOUBLE\_PRECISION} for double precision.
\end{description}
On a new system, it is definitely worth experimenting with single
versus double precision simulations to determine which is faster.
Note that on many current processors (e.g., Intel, AMD, IBM Power),
single precision calculations are often significantly faster; the
difference can typically be 10\% to 25\%. It is therefore often worth
using single precision if you can. We recommend running the same calculation
once in single precision and in double precision on your system and
comparing the seismograms. If they are identical, you should probably
select single precision for your future runs.

@@

The first step in running a spectral-element method computation consists of constructing a high-quality mesh for the region under consideration. We provide two possibilities to do so: (1) relying on an external, hexahedral mesher CUBIT or (2) using the provided, internal mesher \texttt{xmeshfem3D}. In the following, we explain these two approaches.

@@

The basic steps in creating a load-balanced, partitioned mesh with CUBIT are:
\begin{description}
\item [1.] setting up a hexahedral mesh with CUBIT,
\item [2.] exporting the CUBIT mesh into a specfem3d file format and
\item [3.] partitioning the specfem3d mesh files for a chosen number of cores.
\end{description}

Examples are provided in the SPECFEM3D package in the subdirectory \texttt{examples/}. Please feel free to contribute your own example to this package by contacting the CIG Computational Seismology Mailing List  \url{cig-seismo@geodynamics.org}.

@@

For the installation and handling of the CUBIT meshing tool suite, please refer to the CUBIT user manual and documentation.
In order to give you a basic understanding of how to use CUBIT for our purposes, examples are provided in the SPECFEM3D package in the subdirectory \texttt{examples/}. For your convenience, the following examples are provided in the subdirectory \texttt{examples/}:
\begin{description}
\item [{\texttt{homogeneous\_halfspace}}] Basic example creates a single block model and assigns elastic material parameters.
\item [{\texttt{layered\_halfspace}}]  Combines two different, elastic material volumes and creates a refinement layer between the two. This example can be compared for validation against the solutions provided in subdirectory ~\\
\texttt{VALIDATION\_3D\_SEM\_SIMPLER\_LAYER\_SOURCE\_DEPTH/}.
\item [{\texttt{waterlayered\_halfspace}}] Combines an acoustic and elastic material volume as in a schematic marine survey example.
\item [{\texttt{tomographic\_model}}] Creates a single block model whose material properties will have to be read in from a tomographic model file during the databases creation by \texttt{xgenerate\_databases}.
\end{description}

@@

\caption{Screenshots of the CUBIT examples provided in subdirectory  \texttt{examples/}: homogeneous halfspace (top-left), layered halfspace (top-right), waterlayered halfspace (bottom-left) and tomographic model (bottom-right).}

@@

Once the geometric model volumes in CUBIT are meshed, you prepare the model for the exportation with the definition of the material blocks and of the boundary surfaces. Thus, prior to exporting the mesh, you need to define blocks specifying the materials and absorbing boundaries in CUBIT.
This process could be done automatically using the script \texttt{boundary\_definition.py} if the mesh meets some conditions or manually, following the block convention:

@@

For an elastic material:
\begin{description}
\item [material\_id] An integer value which is unique for this material.
\item [Vp] P-Velocity of the material (given in m/s).
\item [Vs] S-Velocity of the material (given in m/s).
\item [rho] density of the material (given in kg/$m^3$).
\item [Q] quality factor to use in case of a simulation with attenuation turned on. It should be between 1 and 9000. In case no attenuation information is available, it can be set to zero. Please note that your Vp- and Vs-velocities are given for a reference frequency. To change this reference frequency, you change the value of
\texttt{ATTENUATION\_f0\_REFERENCE} in the main constants file \texttt{constants.h} found in subdirectory \texttt{src/shared/}.
\item [anisotropic\_flag] Flag describing the anisotropic model to use in case an anisotropic simulation should be conducted. See the file \texttt{model\_aniso.f90} in subdirectory \texttt{src/generate\_databases/} for an implementation of the anisotropic models. In case no anisotropy is available, it can be set to zero.
\end{description}
Note that this material block has to be defined using all the volumes which belong to this elastic material. For volumes belonging to another, different material, you will need to define a new material block.

For an acoustic material:
\begin{description}
\item [material\_id] An integer value which is unique for this material.
\item [Vp] P-Velocity of the material (given in m/s).
\item [0] S-Velocity of the material is ignored.
\item [rho] density of the material (given in kg/$m^3$).
\end{description}

@@

\begin{description}
\item [nummaterial\_velocity\_file] Defines the material properties. For fully defined materials, the formats is:
\begin{lyxcode}
\texttt{domain\_ID} \texttt{material\_ID}  \texttt{rho}  \texttt{vp}  \texttt{vs}  \texttt{Q}  \texttt{anisotropy\_flag}
\end{lyxcode}
where
\texttt{domain\_ID} is 1 for acoustic or 2 for elastic materials, \texttt{material\_ID} a unique identifier,  \texttt{rho} the density in $kg \, m^{-3}$,  \texttt{vp} the P-velocity in $m \, s^{-1}$, \texttt{vs}  the S-velocity in $m \, s^{-1}$, \texttt{Q} the quality factor and \texttt{anisotropy\_flag} an identifier for anisotropic models.
For a tomographic model, the material definition format is:
\begin{lyxcode}
\texttt{domain\_ID} \texttt{material\_ID}  \texttt{tomography}  \texttt{name}
\end{lyxcode}
where
\texttt{domain\_ID} is 1 for acoustic or 2 for elastic materials, \texttt{material\_ID} a negative, unique identifier, \texttt{tomography} keyword for tomographic material definition and \texttt{name} the name of the tomography file. The name is not used so far, rather change the filename defined in the file \texttt{model\_tomography.f90} located in the
\texttt{src/generate\_databases/} directory.
\item [materials\_file] Contains the material associations for each element.
\item [nodes\_coords\_file] Contains the point locations in Cartesian coordinates of the mesh element corners.
\item [mesh\_file] Contains the mesh element connectivity.
\item [free\_surface\_file] Contains the free surface connectivity.
\item [absorbing\_surface\_file\_xmax] Contains the surface connectivity of the absorbing boundary surface at the Xmax.
\item [absorbing\_surface\_file\_xmin] Contains the surface connectivity of the absorbing boundary surface at the Xmin.
\item [absorbing\_surface\_file\_ymax] Contains the surface connectivity of the absorbing boundary surface at the Ymax.
\item [absorbing\_surface\_file\_ymin] Contains the surface connectivity of the absorbing boundary surface at the Ymin.
\item [absorbing\_surface\_file\_bottom] Contains the surface connectivity of the absorbing boundary surface at the bottom.
\end{description}

@@

The SPECFEM3D software package performs large scale simulations in a parallel 'Single Process Multiple Data' way.
The spectral element mesh created with CUBIT needs to be distributed on the processors. This partitioning is executed once and for all prior to
the execution of the solver so it is referred to as a static mapping.

An efficient partitioning is important because it leverages the overall running time of the application.
It amounts to balance the number of elements in each slice while minimizing the communication coasts
resulting from the placement of adjacent elements to different processors. \texttt{decompose\_mesh\_SCOTCH} depends on the Scotch library
which provides efficent static mapping, graph and mesh partitioning routines. Scotch is a free software package developed by
Fran\c {c}ois Pellegrini et al. from LaBRI and INRIA in Bordeaux, France, downloadable from the web page \url{https://gforge.inria.fr/projects/scotch/}. It is more recent than METIS, actively maintained and performs better in many cases.
\\

Prior to compiling \texttt{decompose\_mesh\_SCOTCH}, make sure you have correctly specified the path
of the Scotch library with the option \texttt{-{}-with-scotch-path} of the \texttt{configure} script.
For convenience we provide the source code of SCOTCH, which is released open source under the French CeCILL-C version 1 license,
in directory ~\\
\texttt{src/decompose\_mesh\_SCOTCH/scotch\_5.1.10b}. Please refer to file INSTALL.txt in that directory to see how to compile it.
In the future you should be able to find for recent versions at  \url{http://www.labri.fr/perso/pelegrin/scotch/scotch\_en.html} .
\\

@@

Then you are ready to compile,
in the main directory type '\texttt{make decompose\_mesh\_SCOTCH}'.
If all paths and flags have been set correctly, the executable \texttt{bin/xdecompose\_mesh\_SCOTCH} should be produced.

>>>>

\author{$\copyright$ Princeton University (USA) and\\
University of Pau / CNRS / INRIA (France)\\
Version 2.0.0 \\
"Sesame" 
}

\maketitle

\section*{Authors}
The SPECFEM3D package was first developed by Dimitri Komatitsch and Jeroen Tromp at Harvard University, then Caltech and Princeton, USA, starting in 1998, based in part on earlier work by Dimitri Komatitsch and Jean-Pierre Vilotte at IPG in Paris, France from 1994 to 1997. \\

%Older versions of the code were initially developed by Dimitri Komatitsch at Institut de Physique du Globe (France)
%and then by Dimitri Komatitsch and Jeroen Tromp at Harvard University (USA).

Since then it has been developed and maintained by a development team: in alphabetical order,
Emanuele Casarotti,
Min Chen,
Vala Hj\"orleifsd\'ottir,
Emiljana Jorgji,
Jes\'us Labarta,
Sue Kientz,
Dimitri Komatitsch,
Qinya Liu,
Yang Luo,
Nicolas Le Goff,
Pieyre Le Loher,
Alessia Maggi,
Roland Martin,
Dennis McRitchie,
David Mich\'ea,
Daniel Peter,
Brian Savage,
Leif Strand,
Carl Tape,
Jeroen Tromp.\\

\toall{add full list of developers here and update the AUTHORS file accordingly} \\

%The main developers of the \texttt{SPECFEM3D} source code are Dimitri
%Komatitsch, Jeroen Tromp, and Qinya Liu. The following individuals
%(listed in alphabetical order) have also contributed to the development
%or improvement of the source code: Min Chen, Vala Hj\"orleifsd\'ottir,
%Jes\'us Labarta, and Leif Strand. The following individuals (listed in alphabetical
%order) contributed to this manual: Min Chen, Vala Hj\"orleifsd\'ottir,
%Sue Kientz, Dimitri Komatitsch, Qinya Liu, Alessia Maggi,
%Carl Tape, and Jeroen Tromp.

The manual's cover graphic was created
by Santiago Lombeyda from Caltech's Center for Advanced Computing Research (CACR) \url{http://www.cacr.caltech.edu}. \\

\toall{We should also update the Adobe Illustrator / PDF cover page as well at some point} \\

\newpage{}

@@

The software package SPECFEM3D simulates seismic
wave propagation at the local or regional scale
based upon the spectral-element method (SEM).
The SEM is a continuous Galerkin technique, which can easily be made discontinous \citep{BeMaPa94,KoWoHu02,ChCaVi03,Kop06,WiStBuGh10}; it is then a particular case of the discontinuous Galerkin technique \citep{ReHi73,FaRi99,HuHuRa99,CoKaSh00,GiHeWa02,RiWh03,MoRi05,GrScSc06,BeLaPi06,DuKa06,DeSeWh08,WiStBuGh10,DeSe10,EtChViGl10}, with optimized efficiency because of its tensorized basis functions.
In particular, it can accurately handle very distorted mesh elements \citep{OlSe11}.
Note that in most geological models in the context of seismic wave propagation studies
(except for fault dynamic rupture studies)
a discontinous mesh is not needed because material property contrasts are not drastic and thus a continuous formulation is sufficient. For a detailed introduction to the SEM as applied to regional
seismic wave propagation, please consult \citet{KoVi98,KoTr99,ChKoViCaVaFe07,TrKoLi08} and
in particular \citet{KoLiTrSuStSh04}.

Effects due to lateral variations in compressional-wave speed, shear-wave
speed, density, a 3D crustal model, topography and bathymetry are
included. The package can accommodate
full 21-parameter anisotropy (see~\citet{ChTr07}) as well as lateral
variations in attenuation \citep{SaKoTr10}. Adjoint capabilities and finite-frequency
kernel simulations are included \citep{LiTr06,TrKoLi08,FiIgBuKe09}.

%The SPECFEM3D package was first developed by Dimitri Komatitsch and Jeroen Tromp at Harvard University, then Caltech and Princeton, USA, starting in 1998, based in part on earlier work by Dimitri Komatitsch and Jean-Pierre Vilotte at IPG in Paris, France from 1994 to 1997. Since then it has been developed and maintained by a development team: in alphabetical order,
%Emanuele Casarotti,
%Min Chen,
%Vala Hj\"orleifsd\'ottir,
%Emiljana Jorgji,
%Sue Kientz,
%Dimitri Komatitsch,
%Qinya Liu,
%Yang Luo,
%Nicolas Le Goff,
%Pieyre Le Loher,
%Alessia Maggi,
%Roland Martin,
%Dennis McRitchie,
%David Mich\'ea,
%Daniel Peter,
%Brian Savage,
%Leif Strand,
%Carl Tape,
%Jeroen Tromp
%\toall{add full list of developers here and update the AUTHORS file accordingly}.

%\toall{We should also update the Adobe Illustrator / PDF cover page as well at some point}

All SPECFEM3D software is written in Fortran90 with full portability
in mind, and conforms strictly to the Fortran95 standard. It uses
no obsolete or obsolescent features of Fortran77. The package uses
parallel programming based upon the Message Passing Interface (MPI)
\citep{GrLuSk94,Pac97}.

@@

If you use SPECFEM3D for your own research, please cite at least one
of the following articles:

%\cite{TrKoLi08,VaCaSaKoVi99,LeChLiKoHuTr08,LeChKoHuTr09,LeKoHuTr09,KoMiEr09,KoErGoMi10,KoGoErMi10,WiKoScTr04,KoLiTrSuStSh04,ChKoViCaVaFe07,MaKoDi09,KoViCh10,CaKoLaTiMiLeSnTr08,TrKoHjLiZhPeBoMcFrTrHu10,KoRiTr02,KoTr02a,KoTr02b,KoTr99} or \cite{KoVi98}.

%If you work on geophysical applications, you may be interested in citing some of these application articles as well, among others: \cite{WiKoScTr04,JiTsKoTr05,KrJiKoTr06a,KrJiKoTr06b,LeChLiKoHuTr08,LeChKoHuTr09,LeKoHuTr09,ChFaKo04,FaChKo04,RiRiKoTrHe02,GoAmTaCaSmSaMaKo09,TrKo00,SaKoTr10}.

\item [Forward simulations] ~\\
\cite{TrKoLi08,VaCaSaKoVi99,KoMiEr09,KoErGoMi10,KoGoErMi10,ChKoViCaVaFe07,MaKoDi09,KoViCh10,CaKoLaTiMiLeSnTr08,TrKoHjLiZhPeBoMcFrTrHu10,KoRiTr02,KoTr02a,KoTr02b,KoTr99} or \cite{KoVi98}

\item [Adjoint simulations] ~\\
\cite{TrTaLi05,LiTr06,TrKoLi08,LiTr08,TrKoLi08,TrKoHjLiZhPeBoMcFrTrHu10}

\item [Southern California simulations] ~\\
\cite{KoLiTrSuStSh04,KrJiKoTr06a,KrJiKoTr06b}.

@@

\noindent
If you work on geophysical applications, you may be interested in citing some of these application articles as well, among others:

\item [Anisotropy] ~\\
\cite{ChTr07,JiTsKoTr05,ChFaKo04,FaChKo04,RiRiKoTrHe02,TrKo00}.

\item [Attenuation] ~\\
\cite{SaKoTr10,KoTr02a,KoTr99}.

\item [Topography] ~\\
\cite{LeKoHuTr09,LeChKoHuTr09,LeChLiKoHuTr08,GoAmTaCaSmSaMaKo09,WiKoScTr04}.

The corresponding Bib\TeX{} entries may be found
in file \texttt{USER\_MANUAL/bibliography.bib} or in comments at the
beginning of file \texttt{specfem3D.f90}.

@@

The SPECFEM3D software package relies on the Scotch library to partition meshes created with CUBIT.
Note that we use CUBIT to create meshes of hexahedra, but other packages can be used as well, for instance GiD from \url{http://gid.cimne.upc.es} or Gmsh from \url{http://geuz.org/gmsh}. Even mesh creation packages that generate tetrahedra, for instance TetGen from \url{http://tetgen.berlios.de}, can be used because each tetrahedron can then easily be decomposed into four hexahedra as shown in the picture of the TetGen logo at \url{http://tetgen.berlios.de/figs/Delaunay-Voronoi-3D.gif}; while this approach does not generate hexahedra of optimal quality, it can ease mesh creation in some situations and it has been shown that the spectral-element method can very accurately handle distorted mesh elements \citep{OlSe11}.

@@

\begin{description}
\item [{\texttt{F90}}] Path to the Fortran90 compiler.
\item [{\texttt{MPIF90}}] Path to MPI Fortran90.
\item [{\texttt{MPI\_FLAGS}}] Some systems require this flag to link to
MPI libraries.
\item [{\texttt{FLAGS\_CHECK}}] Compiler flag for non-critical subroutines.
\item [{\texttt{FLAGS\_NO\_CHECK}}] Compiler flag for creating fast, production-run
code for critical subroutines.
\end{description}
The configuration script automatically creates for each executable a corresponding
\texttt{Makefile} in the \texttt{src/} subdirectoy.
The \texttt{Makefile} contains a number of suggested entries for various
compilers, e.g., Portland, Intel, Absoft, NAG, and Lahey. The software
has run on a wide variety of compute platforms, e.g., various PC clusters
and machines from Sun, SGI, IBM, Compaq, and NEC. Select the compiler
you wish to use on your system and choose the related optimization
flags. Note that the default flags in the \texttt{Makefile} are undoubtedly
not optimal for your system, so we encourage you to experiment with
these flags and to solicit advice from your systems administrator.
Selecting the right compiler and optimization flags can make a tremendous
difference in terms of performance. We welcome feedback on your experience
with various compilers and flags.

@@

\begin{description}
\item [{\texttt{CUSTOM\_MPI\_TYPE}}] Set to \texttt{MPI\_REAL} for single
precision and \texttt{MPI\_DOUBLE\_PRECISION} for double precision.
\end{description}
On a new system, it is definitely worth experimenting with single
versus double precision simulations to determine which is faster.
Note that on many current processors (e.g., Intel, AMD, IBM Power),
single precision calculations are often significantly faster; the
difference can typically be 10\% to 25\%. It is therefore often worth
using single precision if you can. We recommend running the same calculation
once in single precision and in double precision on your system and
then comparing the seismograms. If they are identical, you should probably
select single precision for your future runs.

@@

The first step in running a spectral-element simulation consists of constructing a high-quality mesh for the region under consideration. We provide two possibilities to do so: (1) relying on the external, hexahedral mesher CUBIT, or (2) using the provided, internal mesher \texttt{xmeshfem3D}. In the following, we explain these two approaches.

@@

The basic steps in creating a load-balanced, partitioned mesh with CUBIT are:
\begin{description}
\item [1.] setting up a hexahedral mesh with CUBIT,
\item [2.] exporting the CUBIT mesh into a SPECFEM3D file format and
\item [3.] partitioning the SPECFEM3D mesh files for a chosen number of cores.
\end{description}

Examples are provided in the SPECFEM3D package in the subdirectory \texttt{examples/}. We strongly encourage you to contribute your own example to this package by contacting the CIG Computational Seismology Mailing List  \url{cig-seismo@geodynamics.org}.

@@

For the installation and handling of the CUBIT meshing tool suite, please refer to the CUBIT user manual and documentation.
In order to give you a basic understanding of how to use CUBIT for our purposes, examples are provided in the SPECFEM3D package in the subdirectory \texttt{examples/}:
\begin{description}
\item [{\texttt{homogeneous\_halfspace}}] Creates a single block model and assigns elastic material parameters.
\item [{\texttt{layered\_halfspace}}]  Combines two different, elastic material volumes and creates a refinement layer between the two. This example can be compared for validation against the solutions provided in subdirectory ~\\
\texttt{VALIDATION\_3D\_SEM\_SIMPLER\_LAYER\_SOURCE\_DEPTH/}.
\item [{\texttt{waterlayered\_halfspace}}] Combines an acoustic and elastic material volume as in a schematic marine survey example.
\item [{\texttt{tomographic\_model}}] Creates a single block model whose material properties will have to be read in from a tomographic model file during the databases creation by \texttt{xgenerate\_databases}.
\end{description}

@@

\caption{Screenshots of the CUBIT examples provided in subdirectory  \texttt{examples/}: homogeneous halfspace (top-left), layered halfspace (top-right), water layered halfspace (bottom-left) and tomographic model (bottom-right).}

@@

Once the geometric model volumes in CUBIT are meshed, you prepare the model for exportation with the definition of material blocks and boundary surfaces. Thus, prior to exporting the mesh, you need to define blocks specifying the materials and absorbing boundaries in CUBIT.
This process could be done automatically using the script \texttt{boundary\_definition.py} if the mesh meets some conditions or manually, following the block convention:

@@

For an elastic material:
\begin{description}
\item [material\_id] An integer value which is unique for this material.
\item [Vp] P-wave speed of the material (given in m/s).
\item [Vs] S-wave speed of the material (given in m/s).
\item [rho] density of the material (given in kg/m$^3$).
\item [Q] quality factor to use in case of a simulation with attenuation turned on. It should be between 1 and 9000. In case no attenuation information is available, it can be set to zero. Please note that your Vp- and Vs-speeds are given for a reference frequency. To change this reference frequency, you change the value of
\texttt{ATTENUATION\_f0\_REFERENCE} in the main constants file \texttt{constants.h} found in subdirectory \texttt{src/shared/}.
\item [anisotropic\_flag] Flag describing the anisotropic model to use in case an anisotropic simulation should be conducted. See the file \texttt{model\_aniso.f90} in subdirectory \texttt{src/generate\_databases/} for an implementation of the anisotropic models. In case no anisotropy is available, it can be set to zero.
\end{description}
Note that this material block has to be defined using all the volumes which belong to this elastic material. For volumes belonging to another, different material, you will need to define a new material block.

For an acoustic material:
\begin{description}
\item [material\_id] An integer value which is unique for this material.
\item [Vp] P-wave speed of the material (given in m/s).
\item [0] S-wave speed of the material is ignored.
\item [rho] density of the material (given in kg/m$^3$).
\end{description}

@@

\begin{description}
\item [nummaterial\_velocity\_file] Defines the material properties. For fully defined materials, the formats is:
\begin{lyxcode}
\texttt{domain\_ID} \texttt{material\_ID}  \texttt{rho}  \texttt{vp}  \texttt{vs}  \texttt{Q}  \texttt{anisotropy\_flag}
\end{lyxcode}
where
\texttt{domain\_ID} is 1 for acoustic or 2 for elastic materials, \texttt{material\_ID} a unique identifier,  \texttt{rho} the density in $kg \, m^{-3}$,  \texttt{vp} the P-wave speed in $m \, s^{-1}$, \texttt{vs}  the S-wave speed in $m \, s^{-1}$, \texttt{Q} the quality factor and \texttt{anisotropy\_flag} an identifier for anisotropic models.
For a tomographic model, the material definition format is:
\begin{lyxcode}
\texttt{domain\_ID} \texttt{material\_ID}  \texttt{tomography}  \texttt{name}
\end{lyxcode}
where
\texttt{domain\_ID} is 1 for acoustic or 2 for elastic materials, \texttt{material\_ID} a negative, unique identifier, \texttt{tomography} keyword for tomographic material definition and \texttt{name} the name of the tomography file. The name is not used so far, rather change the filename defined in the file \texttt{model\_tomography.f90} located in the
\texttt{src/generate\_databases/} directory.
\item [materials\_file] Contains the material associations for each element.
\item [nodes\_coords\_file] Contains the point locations in Cartesian coordinates of the mesh element corners.
\item [mesh\_file] Contains the mesh element connectivity.
\item [free\_surface\_file] Contains the free surface connectivity.
\item [absorbing\_surface\_file\_xmax] Contains the surface connectivity of the absorbing boundary surface at the Xmax.
\item [absorbing\_surface\_file\_xmin] Contains the surface connectivity of the absorbing boundary surface at the Xmin.
\item [absorbing\_surface\_file\_ymax] Contains the surface connectivity of the absorbing boundary surface at the Ymax.
\item [absorbing\_surface\_file\_ymin] Contains the surface connectivity of the absorbing boundary surface at the Ymin.
\item [absorbing\_surface\_file\_bottom] Contains the surface connectivity of the absorbing boundary surface at the bottom.
\end{description}

These mesh files are needed as input files for the partitioner  \texttt{xdecompose\_mesh\_SCOTCH}
to load-balance the mesh. Please see the next section for further details.

\subsubsection*{Checking the mesh quality}
The quality of the mesh may be inspected more precisely based upon the serial code
in the file  ~\\
\texttt{check\_mesh\_quality\_CUBIT\_Abaqus.f90}. 
Running this code is optional because no information needed by the
solver is generated.

Prior to running and compiling this code, you have to export your mesh in CUBIT to an ABAQUS (.inp) format.
You also have to determine a number of parameters of your mesh, such as the number of nodes and number of elements
and modify the header of the \texttt{check\_mesh\_quality\_CUBIT\_Abaqus.f90} source file. Then, 
in the directory \texttt{src/check\_mesh\_quality\_CUBIT\_Abaqus}, type
\begin{lyxcode}
{\small ./compile\_all.csh}{\small \par}
\end{lyxcode}
and use
\begin{lyxcode}
{\small xcheck\_mesh\_quality\_CUBIT\_Abaqus}{\small \par}
\end{lyxcode}
to generate an AVS output file (\texttt{\small AVS\_meshquality.inp}
in AVS UCD format) or OpenDX output file (\texttt{\small DX\_meshquality.}~\\
\texttt{\small dx}) that can be used to investigate mesh quality,
e.g, skewness of elements and a Gnuplot histogram (\texttt{\small mesh\_quality\_}~\\
\texttt{\small histogram.txt}) that can be plotted with gnuplot (type
`\texttt{\small gnuplot plot\_mesh\_quality\_histogram.gnu}'). The
histogram is also printed to the screen. If you want to start designing
your own meshes, this tool is useful for viewing your creations. You
are striving for meshes with elements with `cube-like' dimensions,
e.g., the mesh should contain no very elongated or skewed elements.

@@

The SPECFEM3D software package performs large scale simulations in a parallel 'Single Process Multiple Data' way.
The spectral-element mesh created with CUBIT needs to be distributed on the processors. This partitioning is executed once and for all prior to
the execution of the solver so it is referred to as a static mapping.

An efficient partitioning is important because it leverages the overall running time of the application.
It amounts to balance the number of elements in each slice while minimizing the communication costs
resulting from the placement of adjacent elements on different processors. \texttt{decompose\_mesh\_SCOTCH} depends on the Scotch library
which provides efficent static mapping, graph and mesh partitioning routines. Scotch is a free software package developed by
Fran\c {c}ois Pellegrini et al. from LaBRI and INRIA in Bordeaux, France, downloadable from the web page \url{https://gforge.inria.fr/projects/scotch/}. It is more recent than METIS, actively maintained and performs better in many cases.
\\

Prior to compiling \texttt{decompose\_mesh\_SCOTCH}, make sure you have correctly specified the path
of the Scotch library with the option \texttt{-{}-with-scotch-path} of the \texttt{configure} script.
For convenience we provide the source code of SCOTCH, which is released open source under the French CeCILL-C version 1 license,
in directory ~\\
\texttt{src/decompose\_mesh\_SCOTCH/scotch\_5.1.10b}. Please refer to file INSTALL.txt in that directory to see how to compile it.
In the future you should be able to find more recent versions at  \url{http://www.labri.fr/perso/pelegrin/scotch/scotch\_en.html} .
\\

@@

When you are ready to compile,
in the main directory type '\texttt{make decompose\_mesh\_SCOTCH}'.
If all paths and flags have been set correctly, the executable \texttt{bin/xdecompose\_mesh\_SCOTCH} should be produced.

The partitioning is done in serial for now (in the next release we will provide a parallel version of that code),
the synopsis is:
\begin{lyxcode}
./bin/xdecompose\_mesh\_SCOTCH  nparts  input\_directory output\_directory
\end{lyxcode}
where
\begin{itemize}
     \item \texttt{nparts} is the number of partitions, i.e., the number of cores for the parallel simulations,
     \item \texttt{input\_directory} is the directory which holds all the files generated by the Python script \texttt{cubit2specfem3d.py} explained in the previous Section~\ref{subsec:Exporting-the-Mesh}, e.g. \texttt{MESH/}, and
     \item \texttt{output\_directory} is the directory for the output of this partitioner which stores ACII-format files named like \texttt{proc******\_Database} for each partition. These files will be needed for creating the distributed databases, and have to reside in the directory  \texttt{LOCAL\_PATH} specified in the main \texttt{Par\_file}, e.g. in directory \texttt{in\_out\_files/DATABASES\_MPI}. Please see Chapter~\ref{cha:Creating-Distributed-Databases} for further details.
\end{itemize}
Note that all the files generated by the Python script  \texttt{cubit2specfem3d.py} must be placed in the \texttt{input\_directory} folder before running the program.
